Globally-centered autocovariances in MCMC

### 2020-08-05 Paper/main.tex
Finished Section 2. Pizza time!

Let $F$ be the target distribution with mean $\mu$ defined on the set $\X \subseteq \mathbb{R}^p$ equipped with a countably generated Borel $\sigma$-field $\mathcal{B}(\X)$. Let $\{X_t\}$ be an $F$-Harris ergodic stationary Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$.
% If $\bar{X} = n^{-1} \sum_{t=1}^{n}X_t$ , then with probability 1, $\bar{X} \to \mu$ as $n \to \infty$. \\
The Markov chain is covariance stationary so that
th ACvF at lag $k$ depends only on $k$ and is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_1, X_{1+k})= \mathbb{E}_F \left[(X_1 - \mu)(X_{1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the quality of  sample statistics. The standard estimator is the sample autocovariance matrix at lag $k$
%
\begin{equation} \label{eq:empirical_acf}
    \hat{\Gamma}(k) = \dfrac{1}{n}\sum_{t=1}^{n-\abs{k}} \left(X_t - \bar{X} \right) \left(X_{t+\abs{k}} - \bar{X} \right)^T\,.
\end{equation}
%
% It is a known fact that most of the Markov chains obtained from practical MCMC algorithms are not stationary. Regardless, Equation~\ref{eq:empirical_acf} provides the empirical estimator for ACF at a particular lag for the Markov chain with any initial distribution (\cite{geyer2011introduction}).\\

@@

\begin{figure}[htbp]
    \begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=\linewidth]{plots/gaussian-Targettrace_n1e4.pdf}
   \caption{$n = 10^4$}
   \label{subfig:gauss-trace_1e4}
 \end{subfigure}
 \begin{subfigure}{0.5\textwidth}
   \centering
   \includegraphics[width=\linewidth]{plots/gaussian-Targettrace_n1e5.pdf} 
   \caption{$n = 10^5$}
   \label{subfig:gauss-trace_1e5}
 \end{subfigure}
    \caption{Target density and Density plot for two chains starting at different values for a Gaussian mixture model with mixture probability $= 0.7, \; \mu_1 = -5, \; \mu_2 = 5, \; \sigma_1 = 1, \textrm{ and } \sigma_2 = 0.5$. (a) $n = 10^4$; chains don't interact. (b) $n = 10^5$; chains interact.}
    \label{fig:gaussian-trace}
\end{figure}

@@

 Suppose we have $m$ replications of Markov chains for same target distribution and $\{X_{st}; \; t \in \mathbb{Z}\}$ denote the $s^{th}$ Markov chain. Let the sample mean of $s^{th}$ Markov chain be denoted by $\bar{X}_s$ and the global mean by $\bar{\bar{X}} = \sum_{s = 1}^{m}\bar{X}_s/m$. The global mean provides a better estimate for the expectation value when the number of simulation per chain are not sufficient for the convergence to kick in. The globally-centered autocovariance function (G-ACF) for the $s^{th}$ Markov chain is given by:
%
\[
\hat{\Gamma}_{G,s}(k) = \dfrac{1}{n} \sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T,
\]
and the averaged globally-centered ACF over all $m$ chains is given by 
\[
\hat{\Gamma}_G(k) = \dfrac{1}{m}\sum_{s=1}^m \hat{\Gamma}_{G,s}(k)\;.
\]
If the starting points of these parallel Markov chains are sufficiently dispersed, G-ACF provides more realistic estimation of lag covariances. \cite{priestley1981spectral} shows that the empirical estimator in Equation \ref{eq:empirical_acf} is biased with the main bias term equal to $\abs{k}\Gamma(k)/n$, ignoring the small order terms arising due to estimation of $\mu$. Though, an unbiased estimator with a divisor of $n - \abs{k}$ instead of $n$ exists, the biased estimator is preferred for its lower mean square error for larger relative $k$  (\cite{priestley1981spectral})

% \subsection{Theoretical results} % (fold)
% \label{sub:theoretical_results}

% subsection theoretical_results (end)

The bias for G-ACF is similar to the bias of sample autocovariance estimator in case of single chain except for a few small order terms that shall be later addressed in theorem \ref{th:G-_expec}. Before getting to the bias results for $\hat{\Gamma}_{RAV}(k)$ for any lag $k$, we introduce an additional notation. For $q \geq 1$, we define
\[
\Phi^{(q)} = \sum\limits_{-\infty}^{\infty}\abs{k}^q \|\Gamma(k)\|
\]
We denote $\Phi^{(1)} \textrm{ by } \Phi$.

The proof of the following theorem is available in Appendix~\ref{appendix:bias}.

{\color{blue} Fix the proof for this.}
\begin{theorem} \label{th:G-_expec} Under stationarity,
\[
   \mathbb{E} \left[\hat{\Gamma}_{G,s}(k) \right] = \left(1- \dfrac{|k|}{n}\right) \left(\Gamma(k) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2}\right)  + o \left(n^{-2} \right)\,,
\]
and consequently,
\[
\mathbb{E} \left[\hat{\Gamma}_{G,s}(k) - \hat{\Gamma}(k)_s \right] = -\dfrac{\abs{k}}{n}\Gamma(k) - \dfrac{1}{m}\left(1- \dfrac{|k|}{n}\right)\left(\dfrac{\Sigma}{n} + \dfrac{\Phi}{n^2}\right) + o\left(n^{-2} \right)\,. 
\]
   % \[
   % \mathbb{E} \left[\hat{\Gamma}_{G}(k) \right] = \left(1- \dfrac{|k|}{n}\right)\left(\Gamma(k) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2}\right)  + \dfrac{2(m-1)}{m}\dfrac{\abs{k}}{n^2} \left(\Sigma - \sum_{h=0}^{n-1}\Gamma(h)\right) + o(n^{-2})
   % \]
\end{theorem}

%  We know that $\hat{\Gamma}_s(k)$ is asymptotically unbiased with a $o(1)$ bias term that remains fixed for any $k$. The bias term is given by \cite{priestley1981spectral} as

%  \begin{equation} \label{eq:priestly}
%      \mathbb{E}[\hat{\Gamma}(k)] = \left(1- \dfrac{\abs{k}}{n}\right)\left(\Gamma(k) - \Var{(\bar{X})}
%  \right)
%  \end{equation}

% As a consequence, if $\Var(\bar{X})$ is finite, then $\Var(\bar{X}) \to 0 \textrm{ as } n \to \infty$. We can break $\hat{\Gamma}_{G-}$ into four parts for all $k \geq 1$ as:
%  \[
%  \hat{\Gamma}_{G} =  \dfrac{1}{m}\sum_{s=1}^{m}\hat{\Gamma}_s(k) - \dfrac{1}{mn}\sum_{s=1}^{m}\sum_{t=1}^{|k|}A_{st}^T - \dfrac{1}{mn}\sum_{s=1}^{m}\sum_{t=n-|k|+1}^{n}A_{st} + \dfrac{n-|k|}{mn}\sum_{s=1}^{m}(\bar{X}_s - \bar{\bar{X}})(\bar{X}_s - \bar{\bar{X}})^T\,,
%  \]
% %
% where $A_{st} = (X_{st}-\bar{X}_s)(\bar{X}_s - \bar{\bar{X}})^T$\\
% The expectation value of the first term is given by equation \ref{eq:priestly} and \ref{eq:song} while the other three terms contribute $o(1)$ terms. The complete proof evaluating the expectation values of last three terms can be found in Appendix subsection \ref{appendix:bias}.
% \end{proof}

% \begin{remark} \label{rmrk:exp_G-f_minus_acf}
% Using theorem \ref{th:G_expec} and equation \ref{eq:priestly},
% \[
% \mathbb{E} \left[\hat{\Gamma}_{G}(k) - \hat{\Gamma}(k) \right] = \dfrac{\abs{k}}{n}\left(1 - \dfrac{1}{m}\right)\left(\dfrac{\Sigma}{n} + \dfrac{\Phi}{n^2}\right) + \dfrac{2(m-1)}{m}\dfrac{\abs{k}}{n^2} \left(\Sigma - \sum_{h=0}^{n-1}\Gamma(h)\right) + o(n^{-2})
% \]
% We know that the sample autocovariance $\hat{\Gamma}(k)$ underestimates the truth due to a negative bias of $\abs{k}\Gamma(k)/n$ for finite $n$. However, in practice, it is preferred to overestimate the autocovariance for finite $n$ than to underestimate it in order to have a more realistic idea about the correlation in a Markov chain at a particular lag.  Note that the RHS in the above expression is positive. As a consequence, G-ACF will counter this negative bias and give more reliable estimates for ACF.

% \end{remark}

When $m = 1$, $\hat{\Gamma}_G(k)$ corresponds to $\hat{\Gamma}(k)$, the expectation of which can be found in \cite{priestley1981spectral}. Theorem~\ref{th:G-_expec} implies that the globally-centered autocovariances are asymptotically unbiased and exhibit reduced bias in finite samples compared to locally-centered autocovariances. The consequences of this are particularly pertinent for the diagonals of $\Gamma$. Special interest is also in the lag-0 autocovariance, that is, the variance of the target distribution.

\begin{remark} \label{rem:lag0_expectation}
\[
\mathbb{E} \left[\hat{\Gamma}_{G}(0) \right] = \Gamma(0) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2} + o(n^{-2})
\]

\end{remark}

% \subsection{Autocorrelation} % (fold)
% \label{sub:autocorrelation}

% subsection autocorrelation (end)
%{\color{blue}  Autocovariances will be used implictly in SVE estimators. But they are explicitly important in ACF plots via autocorrelations.

%Notation in the appendix is $\Gamma^{ii}_{G,s}$
%}
Lag-0 autocovariance are particularly useful in calculating the autocorrelations.The variance of $i^{th}$ component given by $\Gamma^{ii}(0)$. For any component $i$, the autocorrelation is
\[
\rho^{ii}(k) = \dfrac{\Gamma^{ii}(k)}{\Gamma^{ii}(0)}\,,
\]
and is instrumental in visualizing the serial correlation in components of the Markov chain. The typical estimate of the autocorrelation is constructed from  locally-centered autocovariance estimates which as was demonstrated in Section~\ref{sec:intro} can significantly underestimate the autocorrelation. Instead, we calculate the globally-centered autocorrelation using the formula
\[
\hat{\rho}_{G,s}^{(i)}(k) = \dfrac{ \hat{\Gamma}^{(i)}_{G,s} (k)}{\hat{\Gamma}^{(i)}_{G,s} (0)}.
\]
Despite using similar centering in both numerator and denominator, the globally-centered autocorrelation is far greater than locally-centered autocorrelation for each component as is evident from Figure~\ref{fig:gauss:acf}.

% \begin{proof}
%  The proof is in Appendix subsection \ref{appendix:bias}.
% \end{proof}

\section{Variance Estimators for multiple Markov chains} \label{sec:variance_est}

Let $g:\chi \longrightarrow \mathbb{R}^d$ be an F-integrable function. We are interested in the expected value
%
\[
\mu_g = \mathbb{E}_F[g(X)] = \int_{\chi}g(x)F(dx)\;.
\]

Let $Y$ denote the random variable $g(X)$ where $X \sim F$. Since we have $m$ chains of $n$ samples each, let us make following notations: (1) $Y_{st} = g(X_{st})$ for $t \in \{1, 2, ..., n\}$ and $s \in \{1, 2, ..., m\}$; (2) $\bar{Y}_s = \sum_{t=1}^{n}g(X_{st})/n$; and (3) $\bar{\bar{Y}} = \sum_{s=1}^{m}\bar{Y}_s/m$. To assess the reliability of our simulations, we are interested in Monte Carlo error, i.e. $\bar{Y_s} - \mu$ for the $s^{th}$ chain. The sampling distribution for Monte Carlo error is available via Markov chain CLT (\cite{jones2004markov}) if there exists a positive definite matrix $\Sigma$ such that
%
{\color{blue} }
\[
\sqrt{n}(\bar{X}_s-\mu) \xrightarrow{d} N(0,\Sigma)
\]

>>>>

Let $F$ be the target distribution with mean $\mu$ defined on the set $\X \subseteq \mathbb{R}^p$ equipped with a countably generated Borel $\sigma$-field $\mathcal{B}(\X)$. For $s = 1, \dots, m$, let $\{X_{st}; \; t \in \mathbb{Z}\}$ be the $s^{th}$ $F$-Harris ergodic stationary Markov chain \citep[see][for definitions]{meyn:twee:2009} employed to learn characteristics about $F$.
% If $\bar{X} = n^{-1} \sum_{t=1}^{n}X_t$ , then with probability 1, $\bar{X} \to \mu$ as $n \to \infty$. \\
The process is covariance stationary so that
the ACvF at lag $k$ depends only on $k$ and is defined as 
%
\[
    \Gamma(k) = \Cov_F(X_{s1}, X_{s\,1+k})= \mathbb{E}_F \left[(X_{s1} - \mu)(X_{s\,1+k} - \mu)^T \right]\,.
\]
Estimating $\Gamma(k)$ is critical to assessing the quality of the sampler and the quality of sample statistics. Let $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ denote the Monte Carlo estimator of $\mu$ from the $s$th chain. The standard estimator is the sample autocovariance matrix at lag $k$ for the $s$th chain is
%
\begin{equation} \label{eq:empirical_acf}
    \hat{\Gamma}_s(k) = \dfrac{1}{n}\sum_{t=1}^{n-\abs{k}} \left(X_{st} - \bar{X}_s \right) \left(X_{s t+\abs{k}} - \bar{X}_s \right)^T\,.
\end{equation}
%
% It is a known fact that most of the Markov chains obtained from practical MCMC algorithms are not stationary. Regardless, Equation~\ref{eq:empirical_acf} provides the empirical estimator for ACF at a particular lag for the Markov chain with any initial distribution (\cite{geyer2011introduction}).\\

@@

\begin{figure}[htbp]
    \begin{subfigure}{.5\textwidth}
   \centering
   \includegraphics[width=\linewidth]{plots/gaussian-Targettrace_n1e4.pdf}
   \caption{$n = 10^4$}
   \label{subfig:gauss-trace_1e4}
 \end{subfigure}
 \begin{subfigure}{0.5\textwidth}
   \centering
   \includegraphics[width=\linewidth]{plots/gaussian-Targettrace_n1e5.pdf} 
   \caption{$n = 10^5$}
   \label{subfig:gauss-trace_1e5}
 \end{subfigure}
    \caption{Target density and trace plots for two chains starting at different values.}
    \label{fig:gaussian-trace}
\end{figure}

@@

Recall that we run $m$ independent $F$-ergodic Markov chains denoted by $\{X_{st}; \; t \in \mathbb{Z}\}$ for the $s^{th}$ chain. Let the sample mean of the $s^{th}$ Markov chain be $\bar{X}_s = n^{-1} \sum_{t=1}^{n} X_{st}$ and the global mean by $\bar{\bar{X}} = m^{-1}\sum_{s = 1}^{m}\bar{X}_s$. The global mean is naturally superior estimator of $\mu$ than $\bar{X}_s$. We define the globally-centered autocovariance function (G-ACvF) estimator for the $s^{th}$ Markov chain as
%
\[
\hat{\Gamma}_{G,s}(k) = \dfrac{1}{n} \sum_{t=1}^{n-k}(X_{st}-\bar{\bar{X}})(X_{s(t+k)}-\bar{\bar{X}})^T,
\]
In the event that all $m$ Markov chains have been run long enough that $\bar{X}_s \approx \bar{\bar{X}}$ are similar, then  $\hat{\Gamma}_{s} (k) \approx \hat{\Gamma}_{G,s}(k)$. However, for shorter runs or for slow mixing chains, $\Gamma(k)$ is more appropriately estimated by $\hat{\Gamma}_{G,s}$ as it utilizes information from all chains and accounts for disparity between estimates of $\mu$. We quantify the gains in  bias  for the G-ACvF estimator below.
% If the starting points of these parallel Markov chains are sufficiently dispersed, G-ACF provides more realistic estimation of lag covariances. \cite{priestley1981spectral} shows that the empirical estimator in Equation \ref{eq:empirical_acf} is biased with the main bias term equal to $\abs{k}\Gamma(k)/n$, ignoring the small order terms arising due to estimation of $\mu$. Though, an unbiased estimator with a divisor of $n - \abs{k}$ instead of $n$ exists, the biased estimator is preferred for its lower mean square error for larger relative $k$  (\cite{priestley1981spectral})
Let 
\[
\Phi^{(q)} = \sum_{ k= -\infty}^{\infty}\abs{k}^q \Gamma(k)\,,
\]
and let $\Phi^{(1)}$ be denoted by $\Phi$.  The proof of the following theorem is available in Appendix~\ref{appendix:bias}.

\begin{theorem} \label{th:G-_expec} Let $\E_F \|X_{11}\|^{2 + \delta} < \infty$ for some $\delta > 0$. Let $P$ be a  polynomially ergodic Markov chain of order $\xi > (2 + \epsilon)/(1 + 2/\delta)$ for some $\epsilon > 0$. Then,
\[
   \mathbb{E}_F\left[\hat{\Gamma}_{G,s}(k) \right] = \left(1- \dfrac{|k|}{n}\right) \left(\Gamma(k) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2}\right)  + o \left(n^{-2} \right)\,,
\]
and consequently,
\[
\mathbb{E}_F\left[\hat{\Gamma}_{G,s}(k) - \hat{\Gamma}_s(k) \right] = -\dfrac{\abs{k}}{n}\Gamma(k) - \dfrac{1}{m}\left(1- \dfrac{|k|}{n}\right)\left(\dfrac{\Sigma}{n} + \dfrac{\Phi}{n^2}\right) + o\left(n^{-2} \right)\,. 
\]
\end{theorem}

\begin{remark}
Polynomial ergodicity and the moment conditions are required to ensure $\Phi$ and $\Sigma$ are finite. The above result can be stated more generally for $\alpha$-mixing processes, but we limit our attention to Markov chains. 
\end{remark}

When $m = 1$, $\hat{\Gamma}_{G,s}(k) = \hat{\Gamma}_{s}(k)
$ the bias results for which can be found in \cite{priestley1981spectral}. Since lag-covariances are typically positive in MCMC, Theorem~\ref{th:G-_expec} implies that the G-ACvF estimators are asymptotically unbiased and exhibit reduced bias in finite samples compared to locally-centered ACvF estimators. The consequences of this are particularly pertinent for the diagonals of $\Gamma(k)$.
 % Special interest is also in the lag-0 autocovariance, that is, the variance of the target distribution. 
\begin{remark} \label{rem:lag0_expectation}
The variance of the target distribution is the lag covariance at lag 0 and is of particular interest. The bias of the estimator reduces by a factor of $m$:
\[
\mathbb{E} \left[\hat{\Gamma}_{G}(0) \right] = \Gamma(0) - \dfrac{\Sigma}{mn} - \dfrac{\Phi}{mn^2} + o(n^{-2})
\]
\end{remark}

A first use of the ACvFs in MCMC is in constructing ACF plots. Let the target variance of $i${th} component be $\Gamma^{ii}(0)$. For any component $i$, the autocorrelation is defined as
\[
\rho^{ii}(k) = \dfrac{\Gamma^{ii}(k)}{\Gamma^{ii}(0)}\,,
\]
and is instrumental in visualizing the serial correlation in components of the Markov chain. The typical estimate of the autocorrelation is constructed from $\hat{\Gamma}_s(k)$. Instead, we advocate for using G-ACvF estimates so that,
\[
\hat{\rho}_{G,s}^{(i)}(k) = \dfrac{ \hat{\Gamma}^{(i)}_{G,s} (k)}{\hat{\Gamma}^{(i)}_{G,s} (0)}.
\]
The globally-centered autocorrelation provides a far more realistic assessment of the correlation structure of marginal components of the chain  as evidenced in Figure~\ref{fig:gauss:acf}.

We end this section by noting that we can obtain an average G-ACvF and G-ACF over all $m$ chains
\[
\hat{\Gamma}_G(k) = \dfrac{1}{m}\sum_{s=1}^m \hat{\Gamma}_{G,s}(k) \quad \text{ and } \quad \hat{\rho}^{(i)}_G(k) = \dfrac{1}{m}\sum_{s=1}^m \hat{\rho}^{(i)}_{G,s}(k)\,.
\]
The averaged estimators present a measure of the overall correlation structure induced by the Markov transition $P$.

\section{Variance estimators for multiple Markov chains} \label{sec:variance_est}

A critical use of autocovariances is in the assessment of Monte Carlo variability of estimators.  Let $g:\X \longrightarrow \mathbb{R}^d$ be an $F$-integrable function so that we are interested in estimating
%
\[
\mu_g = \mathbb{E}_F[g(X)] = \int_{\X}g(x)F(dx)\;.
\]

Set $\{Y_{st}\}_{t \geq 1} = \{g(X_{st})\}_{t \geq 1}$ for all $s = 1, \dots, m$ and set $\bar{Y}_s = n^{-1}\sum_{t=1}^{n}Y_{st}$  and  $\bar{\bar{Y}} = m^{-1}\sum_{s=1}^{m}\bar{Y}_s$.

To assess the reliability of simulations, we are interested in Monte Carlo error, i.e. $\bar{Y_s} - \mu$ for the $s^{th}$ chain. The sampling distribution for Monte Carlo error is available via Markov chain CLT (\cite{jones2004markov}) if there exists a positive definite matrix $\Sigma$ such that
%
{\color{blue} }
\[
\sqrt{n}(\bar{X}_s-\mu) \xrightarrow{d} N(0,\Sigma)
\]
